Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-23 pkg/paper/Paper/draft.Rnw
Tried to finalize data analysis. Lots of style improvements.
Data analysis: R code simplified. Code chunks moved to the relevant
sections throughout the paper (and not clustered at beginning). The
only remaining major point is how to present the results for the EU
series. (Previoulsy, this used slightly wrong data windows and only
empirical moments as opposed to model-based estimates.)

Style:
Enhanced graphics by improving aspect ratio, size (as included in the
paper), and detailed captions.

Tables have been more unified using less lines (and omitting all vertical
lines as many journals don'

\documentclass[nojss]{jss}
\usepackage[latin1]{inputenc}
\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{amsfonts,amstext,amsmath}
\usepackage{rotating}
\usepackage{url}
%% need no \usepackage{Sweave}

\setkeys{Gin}{width=\textwidth}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

<<preliminaries>>=
## packages
if(Sys.getlocale() != "C") Sys.setlocale("LC_TIME", "English")
library("glogis")
library("fxregime")
library("lattice")
library("moments")

## additional functions
cross_country_moments <- function(countries, date=yearmon(c(1990,1995,2000,2010)),
  probs = c(0.25, 0.5, 0.75))
{
ret <- mat.or.vec(9,length(date)) 
for(jj in 1:length(date)) {
  rval <- sapply(countries, function(country) {
    nb <- results[[country]]$nbreaks 
    m <- t(sapply(results[[country]]$model, moments)) 
    bd <- if(nb < 1) { 
      end(results[[country]]$data) 
    } else {
      breakdates(results[[country]]$breakpoints, breaks = nb) 
    }
    m[sum(as.yearmon(date[jj]) > bd) + 1, ] 
  })
  v1 <- apply(rval, 1, quantile, probs = probs)
  ret[1:3,jj] <- v1[,1] 
  ret[4:6,jj] <- v1[,2] 
  ret[7:9,jj] <- v1[,3] 
}
  return(ret)
}

moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))

## For the creation of the tables
## <<tabelle, eval=FALSE>>=
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
## run results4 (with dummy creation for sales periods)
Sweave("results1s.Rnw")
setwd(file.path(root_dir, "Paper"))
}

## query result file names
fnam <- Sys.glob(file.path("..", "Analysis", "*.rda"))
fnam2 <- grep("2", fnam, fixed = TRUE)
if(length(fnam2) > 0) fnam <- fnam[-fnam2]

## erm <- yearmon(c(
##   Austria = 1995,
##   ...,
##   France = 1995 + 1/12,
##   ...))

## create erm (Exchange rate mechanism bzw. exchange rate mechanism 2
## ERM~II, ERM~I (part of european monetary system expired in 1.1.1999 gave way to ERM II
## if never entered ""right?
## http://web.archive.org/web/20080101022356/http://lexikon.meyers.de/meyers/Europ%C3%A4isches_W%C3%A4hrungssystem

erm <- yearmon(c(
	Austria = 1995,
	Belgium = 1979 + 2/12,
	CzechRepubic = NA, 
	Denmark = 1979 + 2/12,
	Estonia = NA,
	Finland = 1996 + 9/12,
	France = 1979 + 2/12,
	Germany = 1979 + 2/12,
	Greece = 1998 + 2/12,
	Hungary = NA,
	Ireland = 1979 + 2/12,
	Italy = 1979 + 2/12,
	Luxembourg = 1979 + 2/12,
	Netherlands = 1979 + 2/12,
	Poland = NA,
	Portugal = 1992 + 3/12,
	Slovakia = NA,
	Slovenia = NA,
	Spain = 1986 + 5/12,
	Sweden = NA ,
	UK = NA
))

## sequentially load data
## old: load(file.path("..", "Analysis", fnam[i]))

for(i in seq_along(fnam)) {
  load(file.path(root_dir, "Analysis", fnam[i]))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}

table2<- vector(length = length(fnam), mode = "list")
names(table2) <- sapply(strsplit(sapply(strsplit(fnam, "/", fixed = TRUE), tail, 1), ".", fixed = TRUE), head, 1)

for(qq in seq_along(fnam)) {
  load(file.path(root_dir, "Analysis", fnam[qq]))
  table2[[qq]] <- x
}

## compute moments
mom <- lapply(results, function(obj) t(sapply(obj$model, moments)))

## LaTeX formatting
format_moments <- function(i, digits = 3) {
  rval <- format(round(mom[[i]], digits = digits), nsmall = digits)
  rval <- cbind("", rownames(rval), matrix(paste("$", rval, "$", sep = ""), ncol = 3), "", "", "")
  rval[1,1] <- if(is.character(i)) i else names(mom)[i]
  rval[1,6] <- if(is.na(results[[i]]$erm)) "--" else format(results[[i]]$erm)
  rval[1,7] <- if(is.na(results[[i]]$erm2)) "--" else format(results[[i]]$erm2)
  rval[1,8] <- if(is.na(results[[i]]$eur)) "--" else format(results[[i]]$eur)
  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}

format_cross_country <- function(crossmom, digits = 3, moment="mean", quantil="lower") {
  rval <- format(round(crossmom, digits = digits), nsmall = digits)
  rval <- cbind(moment, quantil, matrix(paste("$", rval, "$", sep=""), ncol = 4))
  rval[1,1] <- "mean"
  rval[2:3,1] <- ""
  rval[1,2] <- "lower"
  rval[2,2] <- "medium"
  rval[3,2] <- "upper"
  rval[4,1] <- "variance"
  rval[5:6,1] <- ""
  rval[4,2] <- "lower"
  rval[5,2] <- "medium"
  rval[6,2] <- "upper"
  rval[7,1] <- "skewness"
  rval[8:9,1] <- ""
  rval[7,2] <- "lower"
  rval[8,2] <- "medium"
  rval[9,2] <- "upper"
  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}

get_data <- function(i) {
  rval <- do.call("merge",
    lapply(i, function(j) results[[j]]$data))
  colnames(rval) <- names(results)[i]
  rval
}

get_fitted <- function(i) {
  rval <- do.call("merge",
    lapply(i, function(j) fitted(results[[j]]$breakpoints, breaks=results[[j]]$nbreaks, type = "mean")))
  colnames(rval) <- names(results)[i]
  rval
}

get_xyplot <- function(i)
  xyplot(merge(get_data(i), get_fitted(i)),
    screen = rep(names(results)[i], 2),
    col = rep(c(1, 4), each = length(i)))

xyplot.gefp <- function(x, panel = NULL, ylim = NULL, ...)
{
  cval <- maxBB$computeCritval(0.05, ncol(x$process))
  if(is.null(ylim)) ylim <- range(c(range(x$process), 1.1 * cval, -1.1 * cval))

  gefp_plot <- function(x, y, subscripts, groups, panel = panel.xyplot,
    col = 1, type = "p", pch = 20, lty = 1, lwd = 1, ...)
  {
    col <- rep(as.list(col), length = nlevels(groups))
    type <- rep(as.list(type), length = nlevels(groups))
    pch <- rep(as.list(pch), length = nlevels(groups))
    lty <- rep(as.list(lty), length = nlevels(groups))
    lwd <- rep(as.list(lwd), length = nlevels(groups))

    for(g in 1:nlevels(groups)) {
      idx <- levels(groups)[g] == groups[subscripts]
      if (any(idx)) panel(x[idx], y[idx], ...,
        col = col[[g]], type = type[[g]], pch = pch[[g]],
        lty = lty[[g]], lwd = lwd[[g]])
      grid::grid.lines(y = grid::unit(0, "native"))
      grid::grid.lines(y = grid::unit(cval, "native"), gp = grid::gpar(col = 2))
      grid::grid.lines(y = grid::unit(-cval, "native"), gp = grid::gpar(col = 2))
    }
  }
  if(is.null(panel)) panel <- gefp_plot

  xyplot(x$process, panel = panel, ylim = ylim, ...)
}

## data
data("HICP", package = "glogis")
HICP <- HICP[, colnames(HICP) != "EU"]

## to change to shorter names if necessary
names(HICP) <- c("Austria",  "Belgium", "CzechRepublic", "Denmark", "Estonia", 
"Finland", "France",  "Germany", "Greece", "Hungary", "Ireland",  "Italy", "Luxembourg", "Netherlands",
"Poland", "Portugal",  "Slovakia", "Slovenia",  "Spain", "Sweden", "UK")

if(!file.exists(file.path(root_dir, "Analysis", "glmean.rda"))) {
   glmean <- function(x) {
    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
  }

  hicps <- hicps[, colnames(hicps) != "EU"]

  ## rolling GL means
  hicp_glmean <- rollapply(hicps, 12, glmean)

  setwd(file.path(root_dir, "Analysis"))
	save(hicps, hicp_glmean, file = "glmean.rda")
	setwd(file.path(root_dir, "Paper"))

}

@

% all created dynamically
%\graphicspath{{Figures/}}

@@

\author{Thomas Windberger\\Universit\"at Innsbruck \And
        Achim Zeileis\\Universit\"at Innsbruck
}
\Plainauthor{Thomas Windberger, Achim Zeileis}

\Abstract{
%(... text when everything else finished, present innovation of paper...)
}

\Keywords{inflation rate, structural break, EMU, Generalized Logistic Distribution 
}

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstraße 15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
}

@@

@@

Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the euro is far from being settled. The controversial topics range from the question of whether or not the eurozone is indeed an optimal currency area (OCA) all the way to the very survival of the euro in light of the budgetary problems of some of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context, is in the center of an ongoing debate: this regards the question about short-run and steady-state inflation uncertainty (in context to inflation expectations) dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries, which is the topic of an investigation by \citet{palomba}. \\
The way towards the EMU basically consisted of 3 stages: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2001) the transition period, which ended with the introduction of the euro as legal tender. A possible fourth stage is the continuing integration of new member states, mostly former communist countries. Any new state that wants to join the EMU is first obliged to fulfill the Maastricht criteria, which, besides setting rules for government debts and interest rates, also requires the participation in the ERM~II (Exchange Rate Mechanism~II) for two years, while the exchange rates towards the euro are not allowed to cross the nominal band.\\

@@

Given that a country experienced quite volatile inflation rates, its efforts to meet the convergence criteria are likely to lead to an alteration at least in the mean (since this was one of the requirements) and possibly in the volatility of their respective inflation rates as well. \\
However, a deeper inquiry into this problem is in need of a method able to capture changes in inflation dynamics. We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates and thereby upon the effect of the EMU on a number of European countries both within and without the eurozone. The framework developed in this paper makes it possible to test for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. We focus on changes in the first three moments of these inflation rates, which we assume to be distributed according to a generalized logistic distribution (GL). The GL distribution is a suitable one due to its flexibility to capture nonsymmetry and because of its fatter tails. For this purpose, we develop a new estimation approach building on previous research by \citet{z07}. \\
This \fixme{Jesus: sell method better, why this one and not another} framework allows us to assess the evolution of mean, variance and skewness over time. The skewness parameter turns out to be quite important to measure changes in the long run inflation dynamics, as it changes considerably over the time of the analysis. This trend would be all but invisible using a normal distribution.  
Our estimation technique enables the simultaneous estimation of breaks in the mean (which is important in the context of the Maastricht criteria), the variance (which should be low for the price mechanism to be efficient) and the skewness (where a change in skewness is evidence of a shift of the distribution towards a new ``center of gravity'') of the inflation rates of interest. This than can be used to search for answers regarding the question about the effect of the EMU on the inflation dynamics of its member states or more general to investigate the effects of external shocks.\\
The \fixme{Jesus: maybe comment on results here} remainder of this paper is structured as follows: Section \ref{s2} presents the data, Section \ref{s3} presents the model and the estimation techniques used, Section \ref{s4} illustrates our approach using Slovenia as an example, Section \ref{s5} presents the results and Section \ref{s6} concludes.

\clearpage
\section{Data}\label{s2}

We \fixme{Jesus: yearly inflation rates better; don't start with data in paper} employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
As  shown in a paper by \cite{pltm}, the harmonization of the treatment of sales prices led to considerable changes in some countries (Germany, Portugal, Ireland, Luxembourg, Belgium, Italy and Spain); failing to take this into account may lead to considerable distortions in the estimation. We dealt with the inclusion of sales periods\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} and all the other seasonality in the data by using the X--12--ARIMA method for desaisonalization, as described in \cite{X-12-ARIMA} and \cite{Findley+Monsell+Bell:1998}.

\clearpage
\section{Model}\label{s3}

To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, the number and the timing of the break points are estimated. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
The correlation \fixme{Jesus: yes, but a DGP needs to be assumed, it would need to take care of any autororrelation, wouldn't it?} dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation, the use of alternative tests, as described in \cite{andreou}, is also hindered by the low number of observations, due to the monthly frequencies.\footnote{In some years the Billion Prices Project of the \cite{bpp}, which provides a much higher data frequency, could solve this problem.}

\subsection{Generalized Logistic Distribution}\label{s31}

@@

\begin{eqnarray}
f(y | \theta, \sigma, \delta) & = & \frac{\frac{\delta}{\sigma} \cdot \exp^{-\frac{y-\theta}{\sigma}}}{(1+\exp^{-\frac{y-\theta}{\sigma}})^{(\delta+1)}} 
\label{eqn1}
\end{eqnarray}

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the distribution simplifies to the logistic distribution, for $\delta<1$ it is skewed to the left and for $\delta$$>$1 it is skewed to the right (see Appendix \ref{glscore} for more details). \\
In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tails, which fit these kind of data somewhat better. \citet{won} develop a maximum likelihood \fixme{Jesus: is this not better than what you do here: answer: no, they use completely other framework} estimator if the underlying distribution is a GL distribution. A similar GL distribution is also used in \citet{tolikas} who analyze extreme risk and value--at--risk in the German stock market, although they do not use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in context with expected inflation. \\
\citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\

@@

against the $H_1$ of changes in at least one of the parameters. We first estimate the parameters of the GL--distribution by means of maximum likelihood $\hat{\phi} ~=~ \underset{\phi}{argmax} \sum_{t=1}^n \log f(y_t|\phi)$. The parameter stability is assessed using the empirical scores $s(y_t|\hat{\phi})$ as measures of model deviation.\footnote{for details see Appendix \ref{glscore} and \citet{z07}}  The empirical fluctuation process $\mathit{efp}(\cdot)$ captures deviations from a hypothesized zero mean over time and is given by:

\begin{eqnarray}
\mathit{efp}(t) ~=~ \hat{V}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s(y_i|\hat{\theta}, \hat{\sigma}, \hat{\delta}) \quad (0 \leq t \leq 1), \nonumber
\label{eqn2}
\end{eqnarray}
%(... jesus: thinks this is incomplete ...) \\
\fixme{Jesus: this seems a bit incomplete} where $\hat{V}$ is some consistent estimator of the variance of the scores, which is used to decorrelate the series as to avoid problems due to serial correlation. Then a functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian bridge:

\begin{eqnarray}
\mathit{efp}(\cdot) ~\overset{d}{\rightarrow}~ W^0(\cdot) \nonumber
\end{eqnarray}

@@

After the $\mathit{efp}(\cdot)$ is computed, it conveys information about departures from the $H_0$ of parameter stability over time. For a precise quantification of the actual deviation from its hypothesized zero--mean path, the main idea of the test statistics presented here is to aggregate the $\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics so that critical values and p values can be derived. In this paper, we employ the Supremum of LM (supLM) test by \citet{andrews}. This test is well suited for single break alternatives and performs better if more than just one of the distribution parameters change.

\clearpage
It is given by:

\begin{eqnarray}
\underset{t \in [0.1, 0.9]}{sup} \frac{\Vert \mathit{efp}(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}

where the appropriate p--values can be computed using:

\begin{eqnarray}
\underset{t \in [0.1, 0.9]}{sup} \frac{\Vert W^0(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}

The supLM test computes a test statistics for all possible change points in a given time frame and rejects the null of no change if the maximum is too high.

\subsection{Breakpoint Estimation}

The dating of structural breaks builds upon the work of \cite{bp03}. If a break is detected via the supLM test, the next step is estimating the number and the timing of these break points. Breakpoints $\tau_1,...,\tau_B$ are estimated via maximization of the full segmented likelihood:

\begin{eqnarray}
\sum_{b=1}^{B+1} \sum_{t=\tau_b + 1}^{\tau_b} \log f(y_t | \phi^{(b)}) \nonumber
\end{eqnarray}

If the test statistics is significant (in this case the supLM test), at least one break is estimated. In principle, the optimal segmentation (i.e. the concrete dates of a break) can be computed once B (the number of breakpoints) is known. However, B needs to be based on the observed data as well. The solution is to compute the timing of the breaks for a number of different values for B and then choose B by means of an information criterion, in this case the LWZ (Liu, Wu and Zidek) as developed in \citet{liu}.
All parameters $\tau_1, \dots, \tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic programming.\footnote{More details for this estimation technique are provided in \cite{z10}.}

\clearpage
\subsection{An Example}\label{s4}

<<echo=FALSE>>=
        x <- results$Slovenia
        spval <- sctest(x$efp, functional = supLM(0.1))$p.value
@

>>>>

\documentclass[nojss]{jss}
\usepackage{amsfonts,amstext,amsmath}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

%% need no \usepackage{Sweave}
\setkeys{Gin}{width=\textwidth}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide, keep.source=TRUE}

@@

\author{Thomas Windberger\\Universit\"at Innsbruck \And
        Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Thomas Windberger, Achim Zeileis}

\Abstract{
  text when everything else finished, present innovation of paper
}

\Keywords{inflation rate, structural break, EMU, generalized logistic distribution}

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  A-6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
}

@@

<<preliminaries>>=
## locale settings
if(Sys.getlocale() != "C") Sys.setlocale("LC_TIME", "English")

## packages
library("glogis")
library("fxregime")

## additional moment method
moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))
@

@@

Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the euro is far from being settled. The controversial topics range from the question of whether or not the eurozone is indeed an optimal currency area (OCA) all the way to the very survival of the euro in light of the budgetary problems of some of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context, is in the center of an ongoing debate: this regards the question about short-run and steady-state inflation uncertainty (in context to inflation expectations) dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries, which is the topic of an investigation by \citet{palomba}.

The way towards the EMU basically consisted of 3 stages: stage I (1990--1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2001) the transition period, which ended with the introduction of the euro as legal tender. A possible fourth stage is the continuing integration of new member states, mostly former communist countries. Any new state that wants to join the EMU is first obliged to fulfill the Maastricht criteria, which, besides setting rules for government debts and interest rates, also requires the participation in the ERM~II (Exchange Rate Mechanism~II) for two years, while the exchange rates towards the euro are not allowed to cross the nominal band.

@@

Given that a country experienced quite volatile inflation rates, its efforts to meet the convergence criteria are likely to lead to an alteration at least in the mean (since this was one of the requirements) and possibly in the volatility of their respective inflation rates as well.

However, a deeper inquiry into this problem is in need of a method able to capture changes in inflation dynamics. We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates and thereby upon the effect of the EMU on a number of European countries both within and without the eurozone. The framework developed in this paper makes it possible to test for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. We focus on changes in the first three moments of these inflation rates, which we assume to be distributed according to a generalized logistic distribution (GL). The GL distribution is a suitable one due to its flexibility to capture nonsymmetry and because of its fatter tails. For this purpose, we develop a new estimation approach building on previous research by \citet{z07}.

This \fixme{Jesus: sell method better, why this one and not another} framework allows us to assess the evolution of mean, variance and skewness over time. The skewness parameter turns out to be quite important to measure changes in the long run inflation dynamics, as it changes considerably over the time of the analysis. This trend would be all but invisible using a normal distribution.  
Our estimation technique enables the simultaneous estimation of breaks in the mean (which is important in the context of the Maastricht criteria), the variance (which should be low for the price mechanism to be efficient) and the skewness (where a change in skewness is evidence of a shift of the distribution towards a new ``center of gravity'') of the inflation rates of interest. This than can be used to search for answers regarding the question about the effect of the EMU on the inflation dynamics of its member states or more general to investigate the effects of external shocks.

The \fixme{Jesus: maybe comment on results here} remainder of this paper is structured as follows: Section~\ref{sec:data} presents the data, Section~\ref{sec:model} presents the model and the estimation techniques used, Section~\ref{sec:slovenia} illustrates our approach using Slovenia as an example, Section~\ref{sec:results} presents the results and Section~\ref{sec:conclusion} concludes.

\clearpage
\section{Data}\label{sec:data}

<<data>>=
data("hicps", package = "glogis")
eu <- na.omit(hicps[, "EU"])
hicps <- hicps[, colnames(hicps) != "EU"]
@

We \fixme{Jesus: yearly inflation rates better; don't start with data in paper}
employ seasonally unadjusted HICP series of 21 countries ranging from January
1990 to December 2010 that are obtained from the \citet{oecd}. Countries
included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,
Finland,	France,	Germany,	Greece,	Hungary,	Ireland,
Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia,
Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as
well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta
are not included since they are very small economies.

The countries in this sample can be divided into three different groups. First
the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and
Spain).  Second, the EU members not participating in the ERM~II: Czech Republic,
Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a
member of the EU and the ERM~II, but not yet a member of the EMU.

As  shown in a paper by \cite{pltm}, the harmonization of the treatment of sales
prices led to considerable changes in some countries (Germany, Portugal,
Ireland, Luxembourg, Belgium, Italy and Spain); failing to take this into
account may lead to considerable distortions in the estimation. We dealt with
the inclusion of sales periods\footnote{This was demanded by a Commission
regulation, details may be found in: \citet{eurlex} and in the
Appendix~\ref{sec:sales}.} and all the other seasonality in the data by using
the X-12-ARIMA method for desaisonalization, as described in \cite{X-12-ARIMA}
and \cite{Findley+Monsell+Bell:1998}.

\clearpage
\section{Model}\label{sec:model}

To monitor changes in the dynamics of the inflation rates, we follow the usual
strategy employed when testing for structural breaks. First, the model is fitted
for the whole sample using the generalized logistic distribution. This then
enables testing for the stability of the parameter estimates. If there is
evidence of significant changes, the number and the timing of the break points
are estimated.

This framework allows to trace the evolution of mean, variance and skewness of
the inflation series and thus interpreting deviations in any of these parameters
as possibly due to a change in monetary policy, is made feasible. Any huge
intervention or change is captured by fluctuations in the parameter stability of
the series and can be quantified and related to a distinct time frame.

The correlation \fixme{Jesus: yes, but a DGP needs to be assumed, it would need
to take care of any autororrelation, wouldn't it?} dynamics of the series is of
less importance to this framework, which concentrates on changes in mean,
variance and skewness and does not focus on autocorrelation, the use of
alternative tests, as described in \cite{andreou}, is also hindered by the low
number of observations, due to the monthly frequencies.\footnote{In some years
the Billion Prices Project of the \cite{bpp}, which provides a much higher data
frequency, could solve this problem.}

\subsection{Generalized logistic distribution}\label{sec:gld}

@@

\begin{eqnarray}
f(y | \theta, \sigma, \delta) & = & \frac{\frac{\delta}{\sigma} \cdot \exp^{-\frac{y-\theta}{\sigma}}}{(1+\exp^{-\frac{y-\theta}{\sigma}})^{(\delta+1)}} 
\label{eq:dglogis}
\end{eqnarray}

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the
distribution simplifies to the logistic distribution, for $\delta<1$ it is
skewed to the left and for $\delta$$>$1 it is skewed to the right (see
Appendix~\ref{sec:gld2} for more details).

In econometrics, the logistic distribution is often used in income distributions
and growth models. This is due to  its fatter tails, which fit these kind of
data somewhat better. \citet{won} develop a maximum likelihood \fixme{Jesus: is
this not better than what you do here: answer: no, they use completely other
framework} estimator if the underlying distribution is a GL distribution. A
similar GL distribution is also used in \citet{tolikas} who analyze extreme risk
and value-at-risk in the German stock market, although they do not use a shape
parameter. Regarding inflation rates, the GL distribution is -- to our best
knowledge -- only used in context with expected inflation.

\citet{batchelor} use a logistic distribution (not its generalization) to model
the distribution of mean expected inflation rates.

@@

against the $H_1$ of changes in at least one of the parameters. We first
estimate the parameters of the GL-distribution by means of maximum likelihood
$\hat{\phi} ~=~ \argmax_\phi \sum_{t=1}^n \log f(y_t|\phi)$. The parameter
stability is assessed using the empirical scores $s(y_t|\hat{\phi})$ as measures
of model deviation.\footnote{for details see Appendix~\ref{sec:gld2} and
\citet{z07}}  The empirical fluctuation process $\mathit{efp}(\cdot)$ captures
deviations from a hypothesized zero mean over time and is given by
%
\begin{eqnarray}
\mathit{efp}(t) ~=~ \hat{V}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s(y_i|\hat{\theta}, \hat{\sigma}, \hat{\delta}) \quad (0 \leq t \leq 1), \nonumber
\label{eq:efp}
\end{eqnarray}
%
where $\hat{V}$ is some consistent estimator of the variance of the scores,
which is used to decorrelate the series as to avoid problems due to serial
correlation. Then a functional central limit theorem (FCLT) for
$\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian
bridge: $\mathit{efp}(\cdot) ~\overset{d}{\rightarrow}~ W^0(\cdot)$.

@@

After the $\mathit{efp}(\cdot)$ is computed, it conveys information about
departures from the $H_0$ of parameter stability over time. For a precise
quantification of the actual deviation from its hypothesized zero-mean path, the
main idea of the test statistics presented here is to aggregate the
$\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics
so that critical values and p values can be derived. In this paper, we employ
the sup$\mathit{LM}$ test of \citet{andrews}. This test is well suited for
single break alternatives and performs better if more than just one of the
distribution parameters change.

It is given by:
%
\begin{eqnarray}
\sup_{t \in [0.1, 0.9]} \frac{\Vert \mathit{efp}(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}
%
where the appropriate $p$~values can be computed using from the corresponding
limiting distribution -- with $\mathit{efp}(t)$ replaced by  $W^0(t)$.
The sup$\mathit{LM}$ test computes a test statistics for all possible change points in a
given time frame and rejects the null of no change if the maximum is too high.

\subsection{Breakpoint estimation}

The dating of structural breaks builds upon the work of \cite{bp03}. If a break
is detected via the sup$\mathit{LM}$ test, the next step is estimating the number and the
timing of these break points. Breakpoints $\tau_1, \dots,\tau_B$ are estimated
via maximization of the full segmented likelihood
%
\begin{eqnarray}
\sum_{b=1}^{B+1} \sum_{t=\tau_b + 1}^{\tau_b} \log f(y_t | \phi^{(b)}) \nonumber
\end{eqnarray}
%
If the test statistics is significant (in this case the sup$\mathit{LM}$ test), at least
one break is estimated. In principle, the optimal segmentation (i.e. the
concrete dates of a break) can be computed once B (the number of breakpoints) is
known. However, B needs to be based on the observed data as well. The solution
is to compute the timing of the breaks for a number of different values for B
and then choose B by means of an information criterion, in this case the LWZ
(Liu, Wu and Zidek) as developed in \citet{liu}. All parameters $\tau_1, \dots,
\tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic
programming.\footnote{More details for this estimation technique are provided in
\cite{z10}.}

<<analysis-results>>=
## run analysis (for seasonally adjusted data)
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
Sweave("results1s.Rnw")
setwd(file.path(root_dir, "Paper"))
}

## query result file names
fnam <- Sys.glob(file.path("..", "Analysis", "*.rda"))
fnam_omit <- c(grep("2", fnam, fixed = TRUE), grep("glmean", fnam, fixed = TRUE))
if(length(fnam_omit) > 0) fnam <- fnam[-fnam_omit]

## Exchange Rate Mechanism information
## ERM~II, ERM~I (part of european monetary system expired in 1999-01-01 gave way to ERM II
## if never entered ""right?
## http://web.archive.org/web/20080101022356/http://lexikon.meyers.de/meyers/Europ%C3%A4isches_W%C3%A4hrungssystem
erm <- yearmon(c(
  Austria = 1995,
  Belgium = 1979 + 2/12,
  CzechRepubic = NA, 
  Denmark = 1979 + 2/12,
  Estonia = NA,
  Finland = 1996 + 9/12,
  France = 1979 + 2/12,
  Germany = 1979 + 2/12,
  Greece = 1998 + 2/12,
  Hungary = NA,
  Ireland = 1979 + 2/12,
  Italy = 1979 + 2/12,
  Luxembourg = 1979 + 2/12,
  Netherlands = 1979 + 2/12,
  Poland = NA,
  Portugal = 1992 + 3/12,
  Slovakia = NA,
  Slovenia = NA,
  Spain = 1986 + 5/12,
  Sweden = NA ,
  UK = NA
))

for(i in seq_along(fnam)) {
  load(file.path(root_dir, "Analysis", fnam[i]))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}
@
